Topics in Algorithms

### 2018-01-14 project/paper.tex
asdasd

The query time reduction is made possible by observing that binary search can be performed on the iteration sequence of the $dist_k(u)$ algorithm described in \cite{Thorup2005}. This is done by precomputing a secondary data structures $\mathcal{T}_u$ for each $u \in V$ in $O(k)$ time. Each $\mathcal{T}_u$ contains information about which subsequences of $0, \hdots, k-1$ that make the modified $dist_k(u,i)$ algorithm (see Alg. \ref{alg:distk}) terminate while maintaining a  stretch of $(2k-1)$. Such subsequences are called $(u,v)$-feasible. Specifically, the subsequences in $T_u$ are \textit{canonical}. $(u,v)$-feasibility and canonicity is defined and discussed in 3.

@@

\section{Oracle with $O(\lg k)$ query time}
As aforementioned, binary search on the iteration sequence of the Thorup-Zwick oracle is performed to drive down the query time to $\lg k$. Starting with the iteration sequence $\mathcal{I}=1,\hdots,k-1$, the idea is to identify $r=O(\lg k)$ subsequences $(\mathcal{I}_1=\mathcal{I}) \supset \mathcal{I}_1 \supset \hdots \supset \mathcal{I}_r$, where for $2 \leq j \leq t$ we have $|\mathcal{I}_{j-1}| \geq 2|\mathcal{I}_j|$. We wish to maintain the invariant that $dist_k(u,v,i)$ applied to the first element of any subsequence $\mathcal{I}_j$ outputs the stretch $(2k-1)$ estimate in $|\mathcal{I}_j|$ time.
\\
\begin{defi}
For vertices $u,v \in V$ an index $j \in \mathcal{I}$ is $(u,v)$-terminal if
\begin{enumerate}
\item $j = k-1$ or
\item $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ 
\end{enumerate}
We note that $j$ attains a $(u,v)$ terminal index, the algorithm terminates in the current or next iteration. \label{def:uvterm}
\end{defi}
The definition of $(u,v)$-terminality captures the essence of when $dist_k(u,v)$ terminates, i.e. when $p_j(u) \in B_v$. This happens either when 1) $j=k-1$, in which case, trivially, $p_{k-1}(u) \in B_v$ or 2) when $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ (in this case $dist_k(u,v)$ terminates either in iteration $j$ or $j+1$), see Alg. \ref{alg:distk}.
\\
\begin{defi}
A subsequence $\mathcal{I'}=i_1, \hdots, i_2$ of $\mathcal{I}$ is $(u,v)$-feasible if
\begin{enumerate}
\item $i_1$ is even and
\item $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$ and
\item $i_2$ is $(u,v)$-terminal
\end{enumerate}
\label{def:uvfeas}
\end{defi}
Condition $1$ and $3$ of $(u,v)$-feasibility employs a symmetric view of iterations of $dist_k(u,v,i)$ in which $u$ and $v$ successively exhanges roles for even an uneven $j$. Condition $2$ ensures that $dist_k(u,v,i)$ returns a $(2k-1)$-approximation when run with the first $(u,v)$-feasible subsequence, as seen in the following lemma:
\\
\begin{lem}
Let $i_1, \hdots, i_2$ be a $(u,v)$-feasible subsequence. Then $dist_k(u,v,i_1)$ gives a $2k-1$ approximation of $d_G(u,v)$ in time $O(i_2-i_1)$.
\paragraph{Proof} By assumption $i_2$ is $(u,v)$-terminal. Therefore, $p_{i_2}(u) \in B_v$ or $p_{i_2 + 1}(v) \in B_u$. Thus $dist_k(u,v,i_1)$ terminates when $j=O(i_2)$ in time $O(i_2-i_1)$.

@@

What lemmas \ref{lem:uvfeasterminates}, \ref{lem:basecase} and \ref{lem:uvfeasibility} entail is that, starting with the $(u,v)$-feasible sequence $\mathcal{I}=0,\hdots,k-1$, we may split it into two subsequences, one in which $(u,v)$-feasibility must be preserved. This splitting can be done recursively until a subsequence $\mathcal{I_r}$ for which $|\mathcal{I_r}|=O(\lg k)$ is determined. Applying $dist_k(u,v,i)$ to the first element of $\mathcal{I_r}$ produces a $(2k-1)$-approximation of the distance from $u$ to $v$ by $(u,v)$-feasibility in time proportional to $|\mathcal{I}_r|=O(\lg k)$.

\begin{figure}[h]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., i]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[i,...,k-1]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

\begin{thm}
For an integer $k\geq 2$, a $(2k - 1)$-approximate distance oracle of G of size $O(kn^{1+1/k})$ and $O(\log k)$ query time can be constructed in $O\left( \min \left\lbrace kmn^{1+\frac{1}{k}}, \sqrt{k}m + kn^{1+c/k}\right\rbrace\right)$ time for some constant $c$.

@@

The query time is bounded by the number of recursions needed before reaching a $(u,v)-feasible$ subsequence of size at most $\log k$. $bdist_k$ uses a binary search, so that at an arbitraty iteration the subsequence will be split in half into subsequences $i_1,...,i$ and $i,...,i_2$. Only the one of these two subsequences that is $(u,v)$-terminal is then recursed upon. As the number of recursions in this binary search is $O(\log k)$ and $i$ and $j$ are known for all subsequences, the query time is $O(\log k)$.
To bound the time to identify these desired $j$, a simple linear search requires $O(k\log k)$ time. This can be improved to $O(k)$ by constructing of a recursive tree $\mathcal{T}_u$ for each $u$ that represents the recursion of $bdist_k$ on the subsequence $\mathcal{I}'$ consisting of all even numbers in $\mathcal{I}$. A bottom-up seach of $\mathcal{T}_u$ can identify for each canonical subsequence (canonical meaning it is a node in $\mathcal{T}$) $i_1,i_1+2,...,i_2$ an index $j$ in the subsequence that maximizes $\delta_j(u)$ in $O(k)$ time. The bottom-up traversal of $T$ requires checking at most $O(\log k)$ subsequences; consider the subsequence $i_1,..,i_2$ in seperate leaf nodes as seen in Fig. \ref{tree}. The path $P$ (red) starting in $p(l_1)$ and ending in $p(l_2)$ has $O(\log k)$ children, which we denote $X$. The union of $X$ clearly covers the entire subsequence $i_1,i_1+2,...,i_2-2,i_2$, and finding a $j$ that maximizes $\delta_j(u)$ in $X$ therefore gives a $j$ that maximizes $\delta_j(u)$ in the subsequence $i_1,...,i_2$. If $j$ is known for each node in $\mathcal{T}$, it follows that finding the desired $j$ for any canonical subsequences is bound by $O(\log k)$\\

The recursion tree for $bdist_k$ for $u$ will have $\theta(k/\log k)$ children and $\theta(k/\log k)$ non-child nodes. Computing $j$ for each subsequence in $T$ will take $O(\log k)$ time per node, so total preprocessing time for $u$ is $O(k)$. Thus, the preprocessing time for finding $j$ for $G$ is $O(nk)\leq O(km)\leq O(kmn^{1/k})$ as desired.
\end{thm}

\begin{Walgo}[ht]{20pt}
  \DontPrintSemicolon
  \SetKwRepeat{Do}{do}{while}
  \SetKwInOut{Input}{Input}
  \SetKwInOut{Output}{Output}
  \Indm
    \Indp
    \BlankLine
    $w \leftarrow p_i(u); j \leftarrow i$

>>>>

The query time reduction is made possible by observing that binary search can be performed on the index sequence of the $dist_k(u)$ algorithm described in \cite{Thorup2005}. This is done by precomputing a secondary data structures $\mathcal{T}_u$ for each $u \in V$ in $O(k)$ time. Each $\mathcal{T}_u$ contains information about which subsequences of $0, \hdots, k-1$ that make the modified $dist_k(u,i)$ algorithm (see Alg. \ref{alg:distk}) terminate while maintaining a  stretch of $(2k-1)$. Such subsequences are called $(u,v)$-feasible. Specifically, the subsequences in $T_u$ are \textit{canonical}. $(u,v)$-feasibility and canonicity is defined and discussed in 3.

@@

\section{Oracle with $O(\lg k)$ query time}
As aforementioned, binary search on the index sequence of the Thorup-Zwick oracle is performed to drive down the query time to $\lg k$. Starting with the index sequence $\mathcal{I}=1,\hdots,k-1$, the idea is to identify $r=O(\lg k)$ subsequences $(\mathcal{I}_1=\mathcal{I}) \supset \mathcal{I}_1 \supset \hdots \supset \mathcal{I}_r$, where for $2 \leq j \leq t$ we have $|\mathcal{I}_{j-1}| \geq 2|\mathcal{I}_j|$. We wish to maintain the invariant that $dist_k(u,v,i)$ applied to the first element of any subsequence $\mathcal{I}_j$ outputs the stretch $(2k-1)$ estimate in $|\mathcal{I}_j|$ time.
\\
\begin{defi}
For vertices $u,v \in V$ an index $j \in \mathcal{I}$ is $(u,v)$-terminal if
\begin{enumerate}
\item $j = k-1$ or
\item $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ 
\end{enumerate}
We note that $j$ attains a $(u,v)$ terminal index, the algorithm terminates in the current or next iteration. \label{def:uvterm}
\end{defi}
The definition of $(u,v)$-terminality captures the essence of when $dist_k(u,v)$ terminates, i.e. when $p_j(u) \in B_v$. This happens either when 1) $j=k-1$, in which case, trivially, $p_{k-1}(u) \in B_v$ or 2) when $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ (in this case $dist_k(u,v)$ terminates either in iteration $j$ or $j+1$), see Alg. \ref{alg:distk}.
\\
\begin{defi}
A subsequence $\mathcal{I'}=i_1, \hdots, i_2$ of $\mathcal{I}$ is $(u,v)$-feasible if
\begin{enumerate}
\item $i_1$ is even and
\item $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$ and
\item $i_2$ is $(u,v)$-terminal
\end{enumerate}
\label{def:uvfeas}
\end{defi}
Condition $1$ and $3$ of $(u,v)$-feasibility employs a symmetric view of iterations of $dist_k(u,v,i)$ in which $u$ and $v$ successively exhanges roles for even an uneven $j$. Condition $2$ ensures that $dist_k(u,v,i)$ returns a $(2k-1)$-approximation when run with the first $(u,v)$-feasible subsequence, as seen in the following lemma:
\\
\begin{lem}
Let $i_1, \hdots, i_2$ be a $(u,v)$-feasible subsequence. Then $dist_k(u,v,i_1)$ gives a $2k-1$ approximation of $d_G(u,v)$ in time $O(i_2-i_1)$.
\paragraph{Proof} By assumption $i_2$ is $(u,v)$-terminal. Therefore, $p_{i_2}(u) \in B_v$ or $p_{i_2 + 1}(v) \in B_u$. Thus $dist_k(u,v,i_1)$ terminates when $j=O(i_2)$ in time $O(i_2-i_1)$.

@@

What lemmas \ref{lem:uvfeasterminates}, \ref{lem:basecase} and \ref{lem:uvfeasibility} entail is that, starting with the $(u,v)$-feasible sequence $\mathcal{I}=0,\hdots,k-1$, we may split it into two subsequences, one in which $(u,v)$-feasibility must be preserved. This splitting can be done recursively until a subsequence $\mathcal{I}_r$ for which $|\mathcal{I}_r|=O(\lg k)$ is determined. Applying $dist_k(u,v,i)$ to the first element of $\mathcal{I}_r$ produces a $(2k-1)$-approximation of the distance from $u$ to $v$ (by $(u,v)$-feasibility of $\mathcal{I}_r$) in time proportional to $|\mathcal{I}_r|=O(\lg k)$.

The following theorem is the main result of \cite{Wulff2013}:
\\
\begin{thm}
For an integer $k\geq 2$, a $(2k - 1)$-approximate distance oracle of G of size $O(kn^{1+1/k})$ and $O(\log k)$ query time can be constructed in $O\left( \min \left\lbrace kmn^{1+\frac{1}{k}}, \sqrt{k}m + kn^{1+c/k}\right\rbrace\right)$ time for some constant $c$.

@@

The query time is bounded by the number of recursions needed before reaching a $(u,v)-feasible$ subsequence of size at most $\log k$. $bdist_k$ uses binary search, to select and recurse on the $(u,v)$-feasible subsequence according to lemma \ref{lem:uvfeasibility}. As the number of recursions in this binary search is $O(\log k)$ and $i$ and $j$ are known for all subsequences, the query time is $O(\log k)$.

The time to compute $j$ for all canonical subsequences using linear search requires $O(k\log k)$ time (an $O(k)$ search for each level of the recursion tree). This can be improved to $O(k)$ by precomputing a tree $\mathcal{T}_u$ in which each node is associated with a canonical subsequence corresponding to the recursion of $bdist_k$ on $\mathcal{I}$. Since the $bdist_k$ terminates when $|\mathcal{I}'| \leq \lg k$, we can stop the recursion when the subsequence associated with a leaf has size at most $\lg k$. Each node should contain the $j_{\mathcal{I}'}$ maximizing $\delta_{j_{\mathcal{I}'}}(u)$ for its subsequence $\mathcal{I}'$. The algorithm used to construct $\mathcal{T}_u$ determines $j_{\mathcal{I}'}$ for $\mathcal{I}'=i_1,\hdots,i_2$ in a bottom-up fashion by letting it attain the value of the maximum $j_{\mathcal{I_c}}$ for all children $\mathcal{I_c}$ on the path from the parent of the child containing $i_1$ to the parent of the child containing $i_2$, depicted in Fig. \ref{tree}. Such a traversal of $\mathcal{T}_u$ requires checking at most $O(\log k)$ subsequences. The path $P$ (red) starting in $p(l_1)$ and ending in $p(l_2)$ has $O(\log k)$ children, which we denote $X$. The union of $X$ clearly covers the entire subsequence $i_1,i_1+2,...,i_2-2,i_2$, and finding a $j$ that maximizes $\delta_j(u)$ in $X$ therefore gives a $j$ that maximizes $\delta_j(u)$ in the subsequence $i_1,...,i_2$. If $j$ is known for each node in $\mathcal{T}$, it follows that finding the desired $j$ for any canonical subsequences is bound by $O(\log k)$\\

The recursion tree for $bdist_k$ for $u$ will have $\theta(k/\log k)$ children and $\theta(k/\log k)$ non-child nodes. Computing $j$ for each subsequence in $T$ will take $O(\log k)$ time per node, so total preprocessing time for $u$ is $O(k)$. Thus, the preprocessing time for finding $j$ for $G$ is $O(nk)\leq O(km)\leq O(kmn^{1/k})$ as desired.
\end{thm}
\begin{figure}[!hbt]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., i]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[i,...,k-1]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

\newpage

\newpage

\appendix
\section{Algorithms}
\begin{Walgo}[ht]{20pt}
  \DontPrintSemicolon
  \SetKwRepeat{Do}{do}{while}
  \SetKwInOut{Input}{Input}
  \SetKwInOut{Output}{Output}
  \Indm
    \Indp
    \BlankLine
    $w \leftarrow p_i(u); j \leftarrow i$
